Performing and evaluating graph-based clustering with rOpenScPCA

### 2024-11-11 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
WIP: much progress. Code basically complete and most text complete

- Perform clustering on an SCE object or Seurat object
- Add clustering results to the object
- Evaluate clustering results using several metrics including:
  - Silhouette width
  - Neighborhood purity
  - Cluster stability, as measured with the Adjusted Rand index

This notebook will use the `SCPCS000001` sample for demonstration and assumes this sample is present in the `OpenScPCA-analysis/data/current` directory.
Please [see this documentation](https://openscpca.readthedocs.io/en/latest/getting-started/accessing-resources/getting-access-to-data/) for more information about downloading OpenScPCA data.

@@

#### Paths

@@

#### Input and output files

Set paths to input and output directories and files in the chunk below.

```{r paths}
# Path to processed SCE file for sample SCPCS000001
input_sce_file <- file.path(data_dir, "SCPCP000001", "SCPCS000001", "SCPCL000001_processed.rds")

# Path to output file where clustering results will be saved
cluster_tsv <- file.path(module_base, "results", "SCPCS000001-cluster-results.tsv")
```

@@

```{r readrds}
sce <- readRDS(input_sce_file)

# Convert to Seurat object for demonstration
seurat_obj <- as.Seurat(
  sce,
  counts = "counts",
  data = "logcounts"
)
```

@@

### Clustering from an SCE object

@@

We also strongly recommend providing a random sed with the `seed` argument to ensure reproducibility.

@@

## Clustering a Seurat object

```{r cluster seurat}
cluster_results <- calculate_clusters(
  seurat_obj,
  seed = params$seed,
  # Since we converted this from an SCE object, its PC name is `"PCA"`
  pc_name = "PCA"
)

@@

For this section, we'll just use the SCE object.

@@

We'll use the function `rOpenScPCA::calculate_silhouette` to calculate the silhouette width, and then visualize the results.
This function assumes columns called `cell_id` and `cluster`.

```{r silhouette}
silhouette_results <- calculate_silhouette(
  sce,
  cluster_results,
  cluster_col = "openscpca_cluster"
)

@@

```{r plot silhouette densit}
ave_silhouette_width <- mean(silhouette_results$silhouette_width)
ggplot(silhouette_results) +
  aes(x = silhouette_width) +
  geom_density(fill = "lightblue") +
  geom_vline(xintercept = ave_silhouette_width, color = "red", size = 1)
```

For most clusters, it is below 0.25, and for cluster 5 it is actually negative.
We also see it is higher for clusters 11 and 12, reaching up to about 0.5.

Overall, these values suggest that clustering is not fantastic and many cells may not well matched. 
We can explore this further by visualizing distributions for each cluster:

```{r plot silhouette violin}
ggplot(silhouette_results) +
  aes(x = cluster, y = silhouette_width) +
  geom_violin() +
  geom_jitter(size = 0.75, alpha = 0.25) +
  stat_summary(color = "red")
```

Cluster 5 looks particularly rough, but cluster 11 seems the most reliable.

@@

We'll use the function `rOpenScPCA::calculate_purity` to calculate the neighborhood purity, and then visualize the results.
This function requires 
This function assumes columns called `cell_id` and `cluster`.

@@

```{r plot purity density}
ave_purity <- mean(purity_results$purity)

ggplot(purity_results) +
  aes(x = purity) +
  geom_density(fill = "lightblue") +
  geom_vline(xintercept = ave_purity, color = "red", size = 1)
```
Not bad! Let's look at individual clusters:

```{r plot purity violin}
ggplot(purity_results) +
  aes(x = cluster, y = purity) +
  geom_violin() +
  geom_jitter(size = 0.75, alpha = 0.25) +
  stat_summary(color = "red")
```
With this measure, clusters 1, 3, and 6 all have average values below 0.75, but the rest are high - in particular clusters 11, 12, and 13.

@@

@@

>>>>

- Perform clustering on an SCE or Seurat object
- Evaluate clustering results using several metrics including:
  - Silhouette width
  - Neighborhood purity
  - Cluster stability, as measured with the Adjusted Rand index
- Save clustering results to an SCE or Seurat object

This notebook will demonstrate use of `rOpenScPCA` with the `SCPCS000001` sample from project `SCPCP000001`, which is assumed present in the `OpenScPCA-analysis/data/current/SCPCP000001` directory.
Please [see this documentation](https://openscpca.readthedocs.io/en/latest/getting-started/accessing-resources/getting-access-to-data/) for more information about obtaining ScPCA data.

@@

### Paths

@@

### Input and output files

```{r paths}
# Path to processed SCE file for sample SCPCS000001
input_sce_file <- file.path(data_dir, "SCPCP000001", "SCPCS000001", "SCPCL000001_processed.rds")
```

@@

```{r read data}
# Read the SCE file
sce <- readRDS(input_sce_file)

# Convert to a Seurat object for demonstration
seurat_obj <- as.Seurat(
  sce,
  counts = "counts",
  data = "logcounts"
)
```

@@

Default use of this function is shown below.

### Clustering on an SCE object

@@

We also strongly recommend providing a random seed with the `seed` argument to ensure reproducibility.

@@

The following chunk shows an example of performing clustering with several non-default parameters:

```{r cluster sce nondefault}
# Calculate clusters with non-default parameters
cluster_results <- calculate_clusters(
  sce,
  algorithm = "leiden",
  nn = 15,
  objective_function = "modularity",
  seed = params$seed
)

### Clustering on a Seurat object

When performing clustering on a Seurat object, `rOpenScPCA::calculate_clusters` assumes that the principal components are stored in a reduced dimension slot named `"pca"`.
If your Seurat has a different name for the principal components, you can provide the name with the argument `pc_name`.
In this example, because we created the Seurat object from the SCE object, its principal components are named `"PCA"`, so we will provide that information.

We also strongly recommend providing a random seed with the `seed` argument to ensure reproducibility.

```{r cluster seurat}
cluster_results <- calculate_clusters(
  seurat_obj,
  # Since we converted this from an SCE object, its PC name is `"PCA"`
  pc_name = "PCA",
  seed = params$seed
)

@@

This section demonstrates how to use several functions for evaluating cluster quality and reliability. 
All examples will use the SCE object for demonstration, but a Seurat object can be provided instead and will provide the same results.
Again, as needed, you may need to specify the `pc_name` argument if the principal components do not use the object type's default name (`"PCA"` for SCE and `"pca"` for Seurat).

@@

We'll use the function `rOpenScPCA::calculate_silhouette` to calculate the silhouette width, and then we'll visualize the results.

This function takes the following arguments:
- An SCE or Seurat object that contains principal components
- A data frame with, at least, columns representing unique cell ids and cluster assignments.
  - By default, these columns should be named `cell_id` and `cluster`, respectively, matching the output of `rOpenScPCA::calculate_clusters`
  - You can override these defaults using the arguments `cell_id_col` and `cluster_col`
- Optionally, `pc_name` if your PCA matrix does not have the default name for the object type (`"PCA"` for SCE and `"pca"` for Seurat)

This function will return the inputted data frame with two additional columns:
- `silhouette_width`: The calculated silhouette width for the cell
- `other`: The closet cluster to the cell besides the cluster to which it belongs, as used in the silhouette width calculation

```{r silhouette}
silhouette_results <- calculate_silhouette(
  sce,
  cluster_results
)

@@

```{r plot silhouette density}
# Calculate mean silhouette width
mean_silhouette_width <- mean(silhouette_results$silhouette_width)

# Visualize values as a density plot, showing the mean value as a vertical line
ggplot(silhouette_results) +
  aes(x = silhouette_width) +
  geom_density(fill = "lightblue") +
  geom_vline(xintercept = mean_silhouette_width, color = "red", size = 1)
```

Looking at the full distribution of silhouette width values, we see most are positive which is encouraging, but the values are mostly below 0.5, and the mean is around 0.2 which is somewhat low.
Silhouette width therefore suggests that, while these clusters do have some structure, they are not very tight.

We can also explore the silhouette width results on a per-cluster basis:

```{r plot silhouette violin}
ggplot(silhouette_results) +
  aes(x = cluster, y = silhouette_width) +
  geom_violin() +
  geom_jitter(size = 0.75, alpha = 0.25) +
  # show the mean in value
  stat_summary(color = "red")
```

This visualization shows us that cluster 5 in particular has the lowest silhouette width values, and clusters 11 and 12 have the highest.

@@

We'll use the function `rOpenScPCA::calculate_purity` to calculate the neighborhood purity, and then visualize the results.

This function takes the following arguments:
- An SCE or Seurat object that contains principal components
- A data frame with, at least, columns representing unique cell ids and cluster assignments.
  - By default, these columns should be named `cell_id` and `cluster`, respectively, matching the output of `rOpenScPCA::calculate_clusters`
  - You can override these defaults using the arguments `cell_id_col` and `cluster_col`
- Optionally, `pc_name` if your PCA matrix does not have the default name for the object type (`"PCA"` for SCE and `"pca"` for Seurat)

This function will return the inputted data frame with two additional columns:
- `purity`: The neighborhood purity for the cell
- `maximum`: The cluster with the highest proportion of observations neighboring the cell

@@

```{r plot purity density}
mean_purity <- mean(purity_results$purity)

# Visualize values as a density plot, showing the mean value as a vertical line
ggplot(purity_results) +
  aes(x = purity) +
  geom_density(fill = "lightblue") +
  geom_vline(xintercept = mean_purity, color = "red", size = 1)
```
This metric shows a relatively high average neighborhood purity, suggesting that neighboring cells tend to belong to the same cluster.

We can again visualize this on a per-cluster basis:

```{r plot purity violin}
ggplot(purity_results) +
  aes(x = cluster, y = purity) +
  geom_violin() +
  geom_jitter(size = 0.75, alpha = 0.25) +
  # show the mean in value
  stat_summary(color = "red")
```

With this view, we see that clusters 1, 3, and 6 all have average values below 0.75, but the rest are higher, in particular clusters 11, 12, and 13.
It's worth noting that both silhouette width and neighborhood purity found that clusters 11 and 12 were among the tightest clusters.

@@

This function takes the following arguments:
- An SCE or Seurat object that contains principal components
- A data frame with, at least, columns representing unique cell ids and cluster assignments.
  - By default, these columns should be named `cell_id` and `cluster`, respectively, matching the output of `rOpenScPCA::calculate_clusters`
  - You can override these defaults using the arguments `cell_id_col` and `cluster_col`
- Optionally, `pc_name` if your PCA matrix does not have the default name for the object type (`"PCA"` for SCE and `"pca"` for Seurat)
TODO UPDATE HERE!!!!!!!!!!!!!!!!!!!!

@@

Since this function uses random sampling when performing bootstrapping, it's important to pass in a random seed for reproducibility.
